\textsf{Are large language models superhuman chemists?}

### 2024-03-28 src/tex/ms.tex
smiret edits

\clearpage
\begin{abstract}
    Large language models (LLMs) have found widespread interest as they can perform tasks they have not been explicitly trained on. 
    This is relevant for the chemical sciences, which face the problem of small and diverse datasets that are frequently in the form of text.
    LLMs have shown promise in addressing these issues and are increasingly being harnessed to predict chemical properties, optimize reactions, and even design and conduct experiments autonomously.

@@

    We curated more than 7,000 question-answer pairs about a wide array of chemical sciences subfields, evaluated leading models, and found that the best models outperformed the best human chemists in our study on average. 
    The models, however, struggle with some chemical reasoning tasks that are easy for human experts and provide overconfident, misleading predictions, such as about chemicals' safety profiles.

@@

\clearpage
\begin{refsection}
\section{Introduction}
\Glspl{llm} are \gls{ml} models trained on massive amounts of text to complete sentences. 
Aggressive scaling of these models has led to a rapid increase in their capabilities,\autocite{brown2020language,zhong2024benchmarking} with the leading models now being able to pass in some evaluations the United States Medical Licensing Examination.\autocite{kung2023performance} 
They also have been shown to design and autonomously perform chemical reactions when augmented with external tools such as web search and synthesis planners.\autocite{openai2024gpt4, Boiko_2023, bran2023chemcrow}
While some see \enquote{sparks of \gls{agi}} in them,\autocite{bubeck2023sparks} others consider them as \enquote{stochastic parrots}---i.e., systems that only regurgitate what they have been trained on.\autocite{bender2021dangers}
In any case, the promise of these models is that they show the ability to solve a wide variety of tasks they have not been explicitly trained on.\autocite{bommasani2021opportunities, anderljung2023frontier}
This led to tremendous economic interest and investment in such generative models, with an expected market of more than \$1.3 trillion (almost \$30 billion for drug discovery applications) by 2032.\autocite{bloomberg}

Chemists and materials scientists have quickly caught on to the mounting attention given to \glspl{llm}, with some voices even suggesting that \enquote{the future of chemistry is language}.\autocite{White_2023}
This statement is motivated by a growing number of reports that use \glspl{llm} to predict properties of molecules or materials,\autocite{jablonka202314, jablonka2024leveraging, xie2024fine, liao2024words, zhang2024chemllm, zhong2024benchmarking}  optimize reactions\autocite{ramos2023bayesian, kristiadi2024sober}  generate materials,\autocite{rubungo2023llm, flam2023language, gruver2024fine} extract information,\autocite{Patiny_2023, Dagdelen_2024, Zheng_2024, lála2023paperqa, caufield2023structured} or to even prototype systems that can autonomously perform reactions in the physical world based on commands provided in natural language.\autocite{bran2023chemcrow, Boiko_2023, darvish2024organa}

In addition, since a lot---if not most---of the information about chemistry is currently stored and communicated in text, there is a strong reason to believe that there is still a lot of untapped potential in \glspl{llm} for chemistry and materials science.
For instance, most insights in chemical research do not directly originate from data stored in databases but rather from the scientists and their ability to interpret data. 
Many of these insights are in the form of text in scientific publications. 
Thus, operating on such texts might be our best way of unlocking these insights and learning from them.
This might ultimately lead to general copilot systems for chemists that can provide answers to questions or even suggest new experiments based on vastly more information than a human could ever read.
Such a usage mode is, in particular, interesting in the face of recent advances in autonomous laboratories.\autocite{Boiko_2023, bran2023chemcrow, darvish2024organa, granda2018controlling, Angello_2022, coley2019robotic, Burger_2020, seifrid2022autonomous}

However, the rapid increase in capabilities of chemical \gls{ml} models led (even before the recent interest in \glspl{llm}) to concerns about the potential for dual use of these technologies, e.g., for the design of chemical weapons.\autocite{gopal2023releasing, ganguli2022red, Urbina_2022, campbell2023censoring, moulange2023towards, urbina2022teachable}
To some extent, this is not surprising as any technology that, for instance, is used to design non-toxic molecules can also be used inversely to predict toxic ones. 
Yet, it is important to realize that while \glspl{llm} can facilitate access to information and tools, this information about controlled chemicals and task-specific tools has been available for years and requires access to controlled facilities such as laboratories.
Still, it is essential to realize that such models' user base is wider than that of chemistry and materials science experts who critically reflect on every output such models produce. 
For example, many students frequently consult these tools---perhaps even to prepare chemical experiments.\autocite{Intelligent.com_2023}
This also applies to users from the general public, who might consider using \glspl{llm} to answer questions about the safety of chemicals.
Thus, for some users, misleading information---especially about safety-related aspects---might lead to harmful outcomes. 
However, even for experts, chemical understanding and reasoning capabilities are essential as they will determine the capabilities and limitations of their models in their work, e.g., in copilot systems for chemists.
Unfortunately, apart from anecdotal reports, there is little evidence on how \glspl{llm} perform compared to expert (human) chemists.

@@

While some benchmarks based on university entrance exams\autocite{Zaki_2024, arora2023llms} or automatic text mining\autocite{song2023honeybee, wei2021chemistryqa} have been proposed, none of them have been widely accepted.
This is likely because they cannot automatically be used with black box (or tool-augmented) systems, do not cover a wide range of topics, or are not carefully validated by experts.
On top of that, the existing benchmarks are not designed to be used with models that support special treatment of molecules or equations and do not provide insights on how the models compare relative to experts.

@@

To contextualize the scores, we also surveyed more than \variable{output/number_experts.txt} experts in chemistry on a subset of the benchmark corpus to be able to compare the performance of current frontier models with (human) chemists of different specializations.
Our results indicate that current frontier models perform \enquote{superhuman} on some aspects of chemistry but, in many cases, including safety-related ones, might be very misleading.
We find that there are still plenty of limitations for current models to overcome, to find use in autonomous systems for chemists. 
Moreover, we conclude that carefully created broad benchmarks represent an important stepping stone for progress in this field.  %and underline the importance of careful evaluation of the capabilities of \glspl{llm} in the chemical sciences.

@@

Importantly, our large pool of questions encompasses a wide range of topics.
This can be seen, for example, in \Cref{fig:topic_barplot} in which we compare the number of questions in different subfields of the chemical sciences (see \Cref{sec:meth-topic} for details on how we assigned topics).
The distribution of topics is also evident from \Cref{fig:question_diversity} in which we visualize the questions in a two-dimensional space using a \gls{pca} on the embeddings of the questions.
In this representation, semantically similar questions are close to each other, and we color the points based on classification into \variable{output/num_topics.txt} topics. 
It is clear that a focus of \chembench (by design) lies on safety-related aspects, which in \Cref{fig:question_diversity} appear as a large distinct cluster.

@@

Since many widely utilized systems only provide access to text completions (and not the raw model outputs), \chembench is designed to operate on text completions.
This is also important with the growing number of tool-augmented systems that are deemed essential for building chemical copilot systems. 
Such systems can augment the capabilities of \glspl{llm} through the use of external tools such as search \glspl{api} or code executors.\autocite{schick2024toolformer, karpas2022mrkl, yao2022react}
In those cases, the \gls{llm} that returns the probabilities for various tokens (that are often used for model evaluations\autocite{Fourrier_Habib_Launay_Wolf}) is only a part of the whole system, and it is not clear how to interpret the probabilities in the context of the whole system.
The text completions, however, are the system's final outputs, which would also be used in a real-world application. 
Hence, we use them for our evaluations.

@@

\paragraph{System performance} 
To understand the current capabilities of \glspl{llm} in the chemical sciences, we evaluated a wide range of leading models\autocite{Huggingface} on the \chembench corpus, including systems augmented with external tools.
An overview of the results of this evaluation is shown in \Cref{fig:human_vs_models_bar}. 
In this figure, we show the percentage of questions that the models answered correctly.
Moreover, we show the worst, best, and average performance of the human experts in our study, which we obtained via a custom web application (\url{chembench.org}) that we used to survey the experts.
Remarkably, the figure shows that the leading \gls{llm}, Claude 3, outperforms the best human in our study in this overall metric and vastly, by more than a factor of two, exceeds the average performance of the experts in our study.
Many other models also outperform the average human performance. 
It is, however, interesting to note that the Galactica model, which was trained specifically for scientific applications, is outperformed by more general models.

@@

In \Cref{fig:confidence_vs_performance}, we show that for some models, there is no significant correlation between the estimated difficulty and whether the models answered the question correctly or not.
For applications in which humans might rely on the models to provide answers, this is a concerning observation highlighting the need for critical reasoning in the interpretation of the model's outputs.\autocite{Li_2023}
For example, for the questions about the safety profile of compounds, GPT-4 reported an average confidence of \variable{output/model_confidence_performance/gpt4_is_pictograms_average_confidence_correct_overall.txt} (on a scale of 1--5) for the \variable{output/model_confidence_performance/gpt4_is_pictograms_num_correct_overall.txt} questions it answered correctly and \variable{output/model_confidence_performance/gpt4_is_pictograms_average_confidence_incorrect_overall.txt} for the \variable{output/model_confidence_performance/gpt4_is_pictograms_num_incorrect_overall.txt} questions it answered incorrectly.
While, on average, the verbalized confidence estimates from Claude 3 seem better calibrated (\Cref{fig:confidence_vs_performance}), they are still misleading for the questions about \gls{ghs} pictograms with an average score of \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_correct_overall.txt} for correct answers and \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_incorrect_overall.txt} for incorrect answers.

>>>>

\clearpage
\begin{abstract}
    Large language models (LLMs) have found widespread interest due to their abilities to process human language and to perform tasks they have not been explicitly trained on. 
    This is relevant for the chemical sciences, which face the problem of small and diverse datasets that are frequently in the form of text.
    LLMs have shown promise in addressing these issues and are increasingly being harnessed to predict chemical properties, optimize reactions, and even design and conduct experiments autonomously.

@@

    We curated more than 7,000 question-answer pairs for a wide array of chemical sciences subfields, evaluated leading evaluated leading open and closed-source LLMs, and found that the best models outperformed the best human chemists in our study on average. 
    The models, however, struggle with some chemical reasoning tasks that are easy for human experts and provide overconfident, misleading predictions, such as about chemicals' safety profiles.

@@

\clearpage
\begin{refsection}
\section{Introduction}
\Glspl{llm} are \gls{ml} models trained on massive amounts of text to complete sentences. 
Aggressive scaling of these models has led to a rapid increase in their capabilities,\autocite{brown2020language,zhong2024benchmarking} with the leading models now being able to pass in some evaluations the United States Medical Licensing Examination.\autocite{kung2023performance} 
They also have been shown to design and autonomously perform chemical reactions when augmented with external tools such as web search and synthesis planners.\autocite{openai2024gpt4, Boiko_2023, bran2023chemcrow}
While some see \enquote{sparks of \gls{agi}} in them,\autocite{bubeck2023sparks} others consider them as \enquote{stochastic parrots}---i.e., systems that only regurgitate what they have been trained on.\autocite{bender2021dangers}
Nevertheless, the promise of these models is that they have show the ability to solve a wide variety of tasks they have not been explicitly trained on.\autocite{bommasani2021opportunities, anderljung2023frontier}
This has led to tremendous economic interest and investment in such generative models, with an expected market of more than \$1.3 trillion (almost \$30 billion for drug discovery applications) by 2032.\autocite{bloomberg}

Chemists and materials scientists have quickly caught on to the mounting attention given to \glspl{llm}, with some voices even suggesting that \enquote{the future of chemistry is language}.\autocite{White_2023}
This statement is motivated by a growing number of reports that use \glspl{llm} to predict properties of molecules or materials,\autocite{jablonka202314, jablonka2024leveraging, xie2024fine, liao2024words, zhang2024chemllm, zhong2024benchmarking}  optimize reactions\autocite{ramos2023bayesian, kristiadi2024sober}  generate materials,\autocite{rubungo2023llm, flam2023language, gruver2024fine} extract information,\autocite{Patiny_2023, Dagdelen_2024, Zheng_2024, lála2023paperqa, caufield2023structured, gupta2022discomat} or to even prototype systems that can autonomously perform reactions in the physical world based on commands provided in natural language.\autocite{bran2023chemcrow, Boiko_2023, darvish2024organa}

In addition, since a lot---if not most---of the information about chemistry is currently stored and communicated in text, there is a strong reason to believe that there is still a lot of untapped potential in \glspl{llm} for chemistry and materials science.\autocite{miret2024llms}
For instance, most insights in chemical research do not directly originate from data stored in databases but rather from the scientists and their ability to interpret data. 
Many of these insights are in the form of text in scientific publications. 
Thus, operating on such texts might be our best way of unlocking these insights and learning from them.
This might ultimately lead to general copilot systems for chemists that can provide answers to questions or even suggest new experiments based on vastly more information than a human could ever read.
Such a usage mode is, in particular, interesting in the face of recent advances in autonomous laboratories.\autocite{Boiko_2023, bran2023chemcrow, darvish2024organa, granda2018controlling, Angello_2022, coley2019robotic, Burger_2020, seifrid2022autonomous}

However, the rapid increase in capabilities of chemical \gls{ml} models led (even before the recent interest in \glspl{llm}) to concerns about the potential for dual use of these technologies, e.g., for the design of chemical weapons.\autocite{gopal2023releasing, ganguli2022red, Urbina_2022, campbell2023censoring, moulange2023towards, urbina2022teachable}
To some extent, this is not surprising as any technology that, for instance, is used to design non-toxic molecules can also be used inversely to predict toxic ones. 
Yet, it is important to recognize that while \glspl{llm} can facilitate access to information and tools, this information about controlled chemicals and task-specific tools has been available for years and requires access to controlled facilities such as laboratories.
Still, it is essential to realize that such models' user base is wider than that of chemistry and materials science experts who critically reflect on every output such models produce. 
For example, many students frequently consult these tools---perhaps even to prepare chemical experiments.\autocite{Intelligent.com_2023}
This also applies to users from the general public, who might consider using \glspl{llm} to answer questions about the safety of chemicals.
Thus, for some users, misleading information---especially about safety-related aspects---might lead to harmful outcomes. 
However, even for experts, chemical understanding and reasoning capabilities are essential as they will determine the capabilities and limitations of their models in their work, e.g., in copilot systems for chemists.
Unfortunately, apart from anecdotal reports, there is little evidence on how \glspl{llm} perform compared to expert (human) chemists.

@@

While some benchmarks based on university entrance exams\autocite{Zaki_2024, arora2023llms} or automatic text mining\autocite{song2023honeybee, wei2021chemistryqa, song-etal-2023-matsci} have been proposed, none of them have been widely accepted.
This is likely because they cannot automatically be used with black box (or tool-augmented) systems, do not cover a wide range of topics, or are not carefully validated by experts.
On top of that, the existing benchmarks are not designed to be used with models that support special treatment of molecules or equations and do not provide insights on how the models compare relative to experts.

@@

To contextualize the scores, we also surveyed more than \variable{output/number_experts.txt} experts in chemistry on a subset of the benchmark corpus to be able to compare the performance of current frontier models with (human) chemists of different specializations.
Our results indicate that current frontier models perform \enquote{superhuman} on some aspects of chemistry but, in many cases, including safety-related ones, might be very misleading.
We find that there are still numerous limitations for current models to overcome, such that they can be directly applied in autonomous systems for chemists.
Moreover, we conclude that carefully created broad benchmarks represent an important stepping stone for progress in this field.  %and underline the importance of careful evaluation of the capabilities of \glspl{llm} in the chemical sciences.

@@

Importantly, our large pool of questions encompasses a wide range of topics.
This can be seen, for example, in \Cref{fig:topic_barplot} in which we compare the number of questions in different subfields of the chemical sciences (see \Cref{sec:meth-topic} for details on how we assigned topics).
The distribution of topics is also evident from \Cref{fig:question_diversity} in which we visualize the questions in a two-dimensional space using a \gls{pca} on the embeddings of the questions.
In this representation, semantically similar questions are close to each other, and we color the points based on classification into \variable{output/num_topics.txt} topics. 
It is clear that a focus of \chembench (by design) lies on safety-related aspects, which in \Cref{fig:question_diversity} appear as a large distinct clusters across the embedding space.

@@

Since many widely utilized systems only provide access to text completions (and not the raw model outputs), \chembench is designed to operate on text completions.
This is also important given the growing number of tool-augmented systems that are deemed essential for building chemical copilot systems.
Such systems can augment the capabilities of \glspl{llm} through the use of external tools such as search \glspl{api} or code executors.\autocite{schick2024toolformer, karpas2022mrkl, yao2022react}
In those cases, the \gls{llm} that returns the probabilities for various tokens (that are often used for model evaluations\autocite{Fourrier_Habib_Launay_Wolf}) is only a part of the whole system, and it is not clear how to interpret the probabilities in the context of the whole system.
The text completions, however, are the system's final outputs, which would also be used in a real-world application. 
Hence, we use them for our evaluations.

@@

\paragraph{System performance} 
To understand the current capabilities of \glspl{llm} in the chemical sciences, we evaluated a wide range of leading models\autocite{Huggingface} on the \chembench corpus, including systems augmented with external tools.
An overview of the results of this evaluation is shown in \Cref{fig:human_vs_models_bar}. 
In this figure, we show the percentage of questions that the models answered correctly.
Moreover, we show the worst, best, and average performance of the human experts in our study, which we obtained via a custom web application (\url{chembench.org}) that we used to survey the experts.
Remarkably, the figure shows that the leading \gls{llm}, Claude 3, outperforms the best human in our study in this overall metric and vastly, by more than a factor of two, exceeds the average performance of the experts in our study.
Many other models also outperform the average human performance. 
Interestingly, the Galactica model, which was trained specifically for scientific applications, underperformed compared to many advanced commercial and some open-source models and barely exceeds the random baseline.

@@

In \Cref{fig:confidence_vs_performance}, we show that for some models, there is no significant correlation between the estimated difficulty and whether the models answered the question correctly or not.
For applications in which humans might rely on the models to provide answers, this is a concerning observation highlighting the need for critical reasoning in the interpretation of the model's outputs.\autocite{Li_2023, miret2024llms}
For example, for the questions about the safety profile of compounds, GPT-4 reported an average confidence of \variable{output/model_confidence_performance/gpt4_is_pictograms_average_confidence_correct_overall.txt} (on a scale of 1--5) for the \variable{output/model_confidence_performance/gpt4_is_pictograms_num_correct_overall.txt} questions it answered correctly and \variable{output/model_confidence_performance/gpt4_is_pictograms_average_confidence_incorrect_overall.txt} for the \variable{output/model_confidence_performance/gpt4_is_pictograms_num_incorrect_overall.txt} questions it answered incorrectly.
While, on average, the verbalized confidence estimates from Claude 3 seem better calibrated (\Cref{fig:confidence_vs_performance}), they are still misleading for the questions about \gls{ghs} pictograms with an average score of \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_correct_overall.txt} for correct answers and \variable{output/model_confidence_performance/claude3_is_pictograms_average_confidence_incorrect_overall.txt} for incorrect answers.
